---
title: "Results"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

`%notin%` <- Negate(`%in%`)

state_pop <- data.frame(read.csv('../cleaned_data/inmates_fm.csv', sep=',') %>%
  dplyr::select(YEAR, STATE, Population) %>%
  dplyr::filter(STATE %notin% c("US", "FE", "ST")) %>%
  dplyr::group_by(YEAR, STATE) %>%
  dplyr::summarise(Pop = sum(Population)))
  
us_pop <- data.frame(read.csv('../cleaned_data/inmates_fm.csv', sep=',') %>%
  dplyr::select(YEAR, STATE, Population) %>%
  dplyr::filter(STATE %notin% c("US", "FE", "ST")) %>%
  dplyr::group_by(YEAR) %>%
  dplyr::summarise(Pop = sum(Population)) %>%
  dplyr::mutate(STATE = "US") %>%
  dplyr::select(YEAR, STATE, Pop))

pop <- rbind(state_pop, us_pop)

race <- read.csv('../cleaned_data/clean_race_original.csv', sep=',') %>%
  mutate_if(is.numeric, funs(ifelse(is.na(.), 0, .))) %>%
  dplyr::mutate(white = WHITEM + WHITEF, 
                black = BLACKM + BLACKF,
                other = AIANM + AIANF + ASIANM + ASIANF + 
                  NHPIM + NHPIF + APIM + APIF + TWORACEM +
                  TWORACEF + ADDRACEM + ADDRACEF + UNKRACEM +
                  UNKRACEF + ifelse(YEAR >= 2000, HISPM + HISPF, 0),
                check = white + black + other - TOTRACEM - TOTRACEF,
                tot = TOTRACEM + TOTRACEF,
                other_alt = tot - white - black) %>%
  dplyr::select(-Population) %>%
  merge(., pop, by=c('YEAR', 'STATE'), all.x=T, all.y=F) %>%
  dplyr::select(YEAR, STATE, STATEID, white, black, other, check, tot, other_alt, Pop)

sex <- read.csv('../cleaned_data/clean_race_original.csv', sep=',') %>%
  mutate_if(is.numeric, funs(ifelse(is.na(.), 0, .))) %>%
  dplyr::mutate(male = TOTRACEM,
                female = TOTRACEF) %>%
  dplyr::select(-Population) %>%
  dplyr::select(YEAR, STATE, STATEID, male, female) %>%
  merge(., pop, by=c('YEAR', 'STATE'), all.x=T, all.y=F)
```

The US federal and state prison population increased sharply from the 1970s to around 2010, decreasing modestly thereafter.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}

state_fed <- read.csv('../cleaned_data/clean_race_original.csv', sep=',') %>%
  mutate_if(is.numeric, funs(ifelse(is.na(.), 0, .))) %>%
  dplyr::mutate(pris_pop = TOTRACEM + TOTRACEF) %>%
  dplyr::group_by(YEAR, STATE) %>%
  dplyr::summarise(total_pop = Population, pris_pop = pris_pop) %>%
  dplyr::ungroup()

ggplot(state_fed %>%
         dplyr::filter(STATE == "US") %>%
         dplyr::select(YEAR, pris_pop)) +
  geom_line(mapping = aes(x=YEAR, y=pris_pop / 1e6)) +
  ylim(0, 2) +
  xlab("Year") +
  ylab("Prison population (millions)")

```

# Demographic Trends

The racial breakdown of the prison population between blacks and whites has stayed fairly even over time (implying over-representation of blacks due to their significantly smaller representation in the overall US population).  Other minorities have increased their share of the prison population, in-line with an increased share of the overall population.  The black prison population overtook the white prison population in the years between the 1984 and 1994 crime bills.  This may have been due to the lowered tolerance of drug crimes, although we were unable to obtain the data necessary to investigate this possibility.  From 1994 to 2010, this trend reversed.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}

race_2 <- race %>%
  dplyr::filter(STATEID == 70) %>%
  dplyr::select(-STATEID) %>%
  dplyr::mutate(White = white / Pop * 10000,
                Black = black / Pop * 10000,
                Other = other_alt / Pop * 10000,
                Total = tot / Pop * 10000) %>%
  dplyr::select(-Pop, -other_alt, -check, -white, -black, -other, -tot)

ggplot(race_2 %>%
         tidyr::gather(key="Race", value="N", -YEAR, -STATE) %>%
         dplyr::mutate(Race = factor(Race, levels = c("White", "Black", "Other", "Total"))), 
       aes(x=YEAR)) +
  geom_line(mapping = aes(y=N, colour=Race), stat="identity") +
  ylab('Prisoners per 10000 capita') +
  xlab('Year') +
  theme(legend.position="right")

# one with percentages by race as pct of whole
ggplot(race_2 %>%
         dplyr::mutate(White = White / Total * 100,
                       Black = Black / Total * 100,
                       Other = Other / Total * 100) %>%
         dplyr::select(-Total) %>%
         tidyr::gather(key="Race", value="N", -YEAR, -STATE) %>%
         dplyr::mutate(Race = factor(Race, levels = c("White", "Black", "Other", "Total"))),
       aes(x=YEAR)) +
  geom_line(mapping = aes(y=N, colour=Race)) +
  ylim(0, 100) +
  ylab('Percent of prison population by race') +
  xlab('Year') +
  theme(legend.position="right")

```

The sex breakout of the prison population remains overwhelmingly male, although their share of the population has decreased slightly over time.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
# one with percetnages by gender as pct of whole

sex_2 <- sex %>%
  dplyr::filter(STATEID == 70) %>%
  dplyr::select(-STATEID)

ggplot(sex_2 %>%
         dplyr::mutate(male = male / Pop * 10000,
                       female = female / Pop * 10000) %>%
         dplyr::select(-Pop) %>%
         tidyr::gather(key="Sex", value="N", -YEAR, -STATE),
       aes(x=YEAR)) +
  geom_line(mapping = aes(y=N, colour=Sex)) +
  ylab("Prison population by sex (per 10000 capita)") +
  xlab("Year") +
  theme(legend.position="right")

ggplot(sex_2 %>%
         dplyr::mutate(tot = male + female,
                       male = male / tot * 100,
                       female = female / tot * 100) %>%
         dplyr::select(-Pop, -tot) %>%
         tidyr::gather(key="Sex", value="N", -YEAR, -STATE),
       aes(x=YEAR)) +
  ylim(0, 100) +
  geom_line(mapping = aes(y=N, colour=Sex)) +
  scale_y_continuous(breaks = seq(0, 100, by=20)) +
  ylab("Percent of prison population by sex") +
  xlab("Year") +
  theme(legend.position="right")

```

Aside from the noted subtle trends, race and sex do not alone significantly explain the drastic increases in the prison population.

# State Trends

Louisiana currently has the highest incarceration rate in the country, followed by Delaware, Oklahoma, Mississippi and Arizona.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.height=8}

state <- state_fed %>%
  dplyr::filter(YEAR %in% c(1978, 2016),
                STATE %notin% c("FE", "ST", "US", "DC")) %>%
  dplyr::mutate(per_cap = pris_pop / total_pop * 10000) %>%
  dplyr::arrange(desc(YEAR), per_cap) %>%
  dplyr::mutate(YEAR = factor(YEAR),
                STATE = factor(STATE, levels = unique(.$STATE)))

ggplot(state, aes(x=per_cap, y=STATE, colour=YEAR)) +
  geom_point(state="identity", position_dodge=3) +
  theme(legend.position="bottom") +
  xlab("Prison population by state (per 10000 capita)") +
  ylab("")

```

The primary driver of the increase in US incarceration rate is the south, with Louisiana, Oklahoma and Mississippi showing the largest increases in per-capita incarceration.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.height=8}

state_2 <- state %>%
  dplyr::select(YEAR, STATE, per_cap) %>%
  tidyr::spread(key=YEAR, value=per_cap) %>%
  dplyr::mutate(pct_change = `2016` - `1978`) %>%
  dplyr::select(STATE, pct_change) %>%
  dplyr::arrange(pct_change) %>%
  dplyr::mutate(STATE = factor(STATE, levels=.$STATE))

ggplot(state_2, aes(x=pct_change, y=STATE)) +
  geom_point() +
  #xlim(c(0, 8)) +
  ylab("") +
  xlab("Increase in per-10000-capita prison population from 1976 to 2016") +
  xlim(c(0, 60))
  
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}

state_mapping <- read.csv('../cleaned_data/state_mapping.tsv', sep='\t')

us_map <- map_data('state') %>%
  left_join(state_2 %>% 
              merge(., state_mapping, by.x='STATE', by.y='Abbrev') %>%
              dplyr::mutate(State = tolower(State)) %>%
              dplyr::select(State, pct_change), by=c('region' = 'State'))

ggplot(us_map, mapping = aes(x=long, y=lat, group=group, fill=pct_change)) +
  geom_polygon(color = "gray90", size = 0.1) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  scale_fill_viridis_c(option = "magma") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  theme(legend.position = "bottom",
        strip.background = element_blank())

```

State dynamics appear to be a significant factor in the increase in prison population, given the major disparities in change of per-capita prison population over time.  While all states have seen considerable increases in per-capita prison population, the south has shown a much larger growth than other regions.

# Private Prison Trends

Private prisons 

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}

#priv <- read.csv('../cleaned_data/clean_data_original.csv') %>%
#  dplyr::select(YEAR, STATE, PVOTHM, PVOTHF) %>%
#  dplyr::mutate_if(is.numeric, funs(ifelse(is.na(.), 0, .))) %>%
#  merge(., state_fed, by=c('YEAR', 'STATE')) %>%
#  dplyr::mutate(pct_priv = (PVOTHM + PVOTHF) / pris_pop) %>%
#  dplyr::filter(YEAR == 2016) %>%
#  full_join(state_mapping %>%
#              dplyr::mutate(State = tolower(State)), 
#            by=c('STATE'='Abbrev'))

priv <- read.csv('../cleaned_data/clean_data_original.csv') %>%
  dplyr::select(YEAR, STATE, PVOTHM, PVOTHF) %>%
  dplyr::mutate_if(is.numeric, funs(ifelse(is.na(.), 0, .))) %>%
  merge(., state_fed, by=c('YEAR', 'STATE')) %>%
  dplyr::mutate(pct_priv = (PVOTHM + PVOTHF) / pris_pop) %>%
  full_join(state_mapping %>%
              dplyr::mutate(State = tolower(State)), 
            by=c('STATE'='Abbrev')) %>%
  dplyr::filter(STATE=='US')

ggplot(priv, aes(x=YEAR, y=pct_priv)) +
  geom_line() +
  xlab("Year") +
  ylab("Private prison population as a percent of total prison population")



#us_map_2 <- map_data('state') %>%
#  full_join(priv, by=c('region'='State'))

#ggplot(us_map_2, mapping = aes(x=long, y=lat, group=group, fill=pct_priv)) +
#  geom_polygon(color = "gray90", size = 0.1) +
#  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
#  scale_fill_viridis_c(option = "magma") +
#  theme(axis.title.x=element_blank(),
#        axis.text.x=element_blank(),
#        axis.ticks.x=element_blank(),
#        axis.title.y=element_blank(),
#        axis.text.y=element_blank(),
#        axis.ticks.y=element_blank()) +
#  theme(legend.position = "bottom",
#        strip.background = element_blank())

```


# Conclusions

It appears that the increases in prison population is driven both by by federal policies, such as the crime bills of 1984 and 1994, as well as state-specific policy.  We expect to have seen a more even increase in per-capita prison populations across states if the primary driver was federal legislation, but the evidence is clear that this is not the case.  However, since all states have seen increases in per-capita prison population, it appears that federal legislation plays a role.
